---
title: "TransLink Delay Propagation вЂ” brms Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = TRUE, cache.lazy = FALSE)
```

# Setup

```{r libs, cache=FALSE}
library(brms)
library(dplyr)
library(ggplot2)
library(lubridate)
library(bayesplot)
library(loo)
library(patchwork)
library(arrow)

# parallel chains
options(mc.cores = parallel::detectCores())
```

# Data

Using the v2 parquet export (route 6641 direction 0, since 2026-05-09).
Anomalous/sparse days are excluded automatically by the data_quality chunk below.

```{r load}
pq_path <- tail(
  sort(list.files("../exports", pattern = "route6641_dir0_.*\\.parquet",
                  full.names = TRUE)),
  1
)
if (length(pq_path) == 0) stop("No route6641_dir0_*.parquet found in exports/")
cat(sprintf("Loading: %s\n", pq_path))

raw <- arrow::read_parquet(pq_path) %>%
  filter(
    !is.na(delay_seconds),
    !is.na(previous_stop_delay),
    !is.na(shape_dist_traveled),
    !is.na(hour),
    !is.na(stop_id),
    !is.na(trip_id),
    abs(delay_seconds)        <= 3600,
    abs(previous_stop_delay)  <= 3600
  ) %>%
  mutate(
    timestamp   = as.POSIXct(timestamp, tz = "UTC"),
    date        = as.Date(timestamp),
    hour        = as.integer(hour),
    dow         = as.integer(wday(timestamp, week_start = 1L)),
    trip_id     = as.character(trip_id),
    stop_id     = as.character(stop_id)
  )

cat("Rows after NA + outlier filter:", nrow(raw), "\n")
cat("Date range:", format(min(raw$date)), "в†’", format(max(raw$date)), "\n")
cat("Unique trips:", n_distinct(raw$trip_id),
    " | Unique stops:", n_distinct(raw$stop_id), "\n")
```

```{r data_quality, cache=FALSE}
cat("=== Data Quality ===\n\n")

n_before_anomaly <- nrow(raw)

# NULL check вЂ” should all be 0 after filtering
model_vars <- c("delay_seconds", "previous_stop_delay", "shape_dist_traveled",
                "hour", "dow", "trip_id", "stop_id")
null_rates <- sapply(model_vars, function(v) mean(is.na(raw[[v]])))
if (any(null_rates > 0)) {
  cat("WARN: Non-zero NULL rates after filtering:\n")
  print(null_rates[null_rates > 0])
} else {
  cat("NULL check: all model variables clean (0% NULL)\n")
}

# Data-driven anomaly day detection (replaces hardcoded date exclusions)
daily_stats <- raw %>%
  group_by(date) %>%
  summarise(n = n(), mean_delay = mean(delay_seconds, na.rm = TRUE), .groups = "drop")

overall_mean <- mean(daily_stats$mean_delay)
overall_sd   <- sd(daily_stats$mean_delay)

anomaly_days <- daily_stats %>%
  filter(n < 0.20 * median(daily_stats$n) |
         abs(mean_delay - overall_mean) > 2 * overall_sd) %>%
  pull(date)

cat(sprintf("Anomaly days excluded: %s\n",
            if (length(anomaly_days) > 0) paste(anomaly_days, collapse = ", ") else "none"))

if (length(anomaly_days) > 0) {
  raw <- raw %>% filter(!date %in% anomaly_days)
  cat(sprintf("Rows: %s в†’ %s  (lost %s)\n",
              format(n_before_anomaly, big.mark = ","),
              format(nrow(raw), big.mark = ","),
              format(n_before_anomaly - nrow(raw), big.mark = ",")))
}

# Delay distribution summary
delay_mean <- mean(raw$delay_seconds)
delay_sd   <- sd(raw$delay_seconds)
skew       <- mean(((raw$delay_seconds - delay_mean) / delay_sd)^3)
cat(sprintf("\nDelay distribution: mean=%.1fs  SD=%.1fs  skewness=%.2f\n",
            delay_mean, delay_sd, skew))
cat(sprintf("Date range: %s в†’ %s  (%d unique dates)\n",
            min(raw$date), max(raw$date), n_distinct(raw$date)))
```

## Temporal split

Train on the first 75 % of dates, hold out the last 25 %.
This respects causal time вЂ” the model never sees future delays during training.

```{r split, cache=FALSE}
dates      <- sort(unique(raw$date))
cutoff     <- dates[floor(0.75 * length(dates))]
train_df   <- filter(raw, date <= cutoff)
test_df    <- filter(raw, date >  cutoff)

cat("Train:", nrow(train_df), "rows up to",   format(cutoff), "\n")
cat("Test: ", nrow(test_df),  "rows from",
    format(min(test_df$date)), "в†’", format(max(test_df$date)), "\n")
stopifnot("Test set has < 100 rows вЂ” check data and split" = nrow(test_df) >= 100)
```

---

# Priors

One block shared across all models.
Matches the original project but expressed in brms syntax.
Note: brms' Student-t family (`nu`) defaults to `gamma(2, 0.1)` вЂ” we make
it explicit and add `lb = 2` to match the hard floor in the original Stan code.

```{r priors}
base_priors <- c(
  prior(normal(0, 2),       class = b),        # all regression coefs
  prior(normal(0, 10),      class = Intercept), # global mean
  prior(exponential(1),     class = sigma),     # residual scale
  prior(exponential(1),     class = sd),        # random-effect scales
  prior(gamma(2, 0.1),      class = nu, lb = 2) # Student-t df
)
```

---

# Model 0 вЂ” Baseline

Equivalent to the original `delay_model_hier.stan`:
previous delay + distance + hour (as factor) + dow (cyclic spline) + trip random effect.

## Inspect Stan code first

```{r m0_stancode}
f_m0 <- bf(
  delay_seconds ~ previous_stop_delay + shape_dist_traveled +
    factor(hour) + s(dow, bs = "cc", k = 5) +
    (1 | trip_id)
)

cat(make_stancode(f_m0, data = train_df, family = student(),
                  prior = base_priors,
                  knots = list(dow = c(0.5, 7.5))))
```

## Fit

```{r m0_fit}
m0 <- brm(
  f_m0,
  data    = train_df,
  family  = student(),
  prior   = base_priors,
  knots   = list(dow = c(0.5, 7.5)),
  iter    = 2000, warmup = 1000, chains = 4, seed = 42,
  control = list(adapt_delta = 0.95),
  file    = "models/brms_m0"
)
```

## Diagnostics

```{r m0_diag, fig.height=3}
# Chain mixing for the parameters we care most about
mcmc_trace(as.array(m0), pars = c("b_Intercept", "b_previous_stop_delay",
                                   "b_shape_dist_traveled", "sigma", "nu"))
```

```{r m0_summary}
summary(m0, pars = c("b_Intercept", "b_previous_stop_delay",
                      "b_shape_dist_traveled", "sigma", "nu"))
```

```{r m0_ppc, fig.height=3}
# Posterior predictive check using Stan's y_rep вЂ” correct multi-draw version
pp_check(m0, ndraws = 50, type = "dens_overlay") +
  coord_cartesian(xlim = c(-600, 600)) +
  labs(title = "M0 PPC: 50 posterior predictive draws vs observed")
```

---

# Model 1 вЂ” Cyclic spline for hour

`factor(hour)` gives 24 independent, unsmoothed parameters.
A cyclic cubic spline (`bs = "cc"`) reduces this to ~8 basis coefficients
and forces the function to be smooth and periodic across midnight (hour 23 в†’ 0).

`knots = list(hour = c(0, 24))` tells mgcv the period is 24 hours.

## Inspect Stan code first

```{r m1_stancode}
f_m1 <- bf(
  delay_seconds ~ previous_stop_delay + shape_dist_traveled +
    s(hour, bs = "cc", k = 8) +
    s(dow, bs = "cc", k = 5) +
    (1 | trip_id)
)

# brm() needs knots passed separately вЂ” inspect using make_stancode + make_standata
stancode_m1 <- make_stancode(f_m1, data = train_df, family = student(),
                              prior = base_priors,
                              knots = list(hour = c(0, 24), dow = c(0.5, 7.5)))
cat(stancode_m1)
```

Notice the spline basis matrix `Zs_` in the generated Stan data вЂ” brms
pre-computes it and passes it as data, not parameters.

## Fit

```{r m1_fit}
m1 <- brm(
  f_m1,
  data    = train_df,
  family  = student(),
  prior   = base_priors,
  knots   = list(hour = c(0, 24), dow = c(0.5, 7.5)),
  iter    = 2000, warmup = 1000, chains = 4, seed = 42,
  control = list(adapt_delta = 0.95),
  file    = "models/brms_m1"
)
```

## Compare hour effects: factor vs spline

```{r m1_hour_compare, fig.height=3}
# M0: extract factor(hour) posterior means (reference level = hour 0, excluded from fixef)
hour_m0 <- as.data.frame(fixef(m0)) %>%
  tibble::rownames_to_column("term") %>%
  filter(grepl("factorhour", term)) %>%
  mutate(hour = as.integer(gsub("factorhour", "", term)))

# M1: dow is now a numeric cyclic spline вЂ” no factor level issues.
# Fix dow to Monday (1) and vary hour across its full range.
hour_grid <- data.frame(
  hour                = 0:23,
  dow                 = 1L,
  previous_stop_delay = 0,
  shape_dist_traveled = mean(train_df$shape_dist_traveled),
  trip_id             = train_df$trip_id[1]
)
f_vals    <- as.data.frame(fitted(m1, newdata = hour_grid,
                                   re_formula = NA, allow_new_levels = TRUE))
hour_m1_df <- dplyr::bind_cols(hour_grid, f_vals)

p_m0_hour <- ggplot(hour_m0, aes(x = hour, y = Estimate)) +
  geom_line(color = "steelblue") +
  geom_ribbon(aes(ymin = `Q2.5`, ymax = `Q97.5`), alpha = 0.2, fill = "steelblue") +
  labs(title = "M0: 24 independent hour effects", x = "Hour", y = "Predicted delay (s)") +
  theme_minimal()

p_m1_hour <- ggplot(hour_m1_df, aes(x = hour, y = Estimate)) +
  geom_line(color = "coral") +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, fill = "coral") +
  labs(title = "M1: cyclic spline (8 basis fns, avg over dow)", x = "Hour", y = "Predicted delay (s)") +
  theme_minimal()

p_m0_hour + p_m1_hour
```

---

# Model 2 вЂ” Extended: stop effects + cyclic hour

Adds a stop-level random effect вЂ” the key improvement over M0 that was shown
to win on LOO in the original project.

## Inspect Stan code first

```{r m2_stancode}
f_m2 <- bf(
  delay_seconds ~ previous_stop_delay + shape_dist_traveled +
    s(hour, bs = "cc", k = 8) +
    s(dow, bs = "cc", k = 5) +
    (1 | trip_id) +
    (1 | stop_id)
)

cat(make_stancode(f_m2, data = train_df, family = student(),
                  prior = base_priors,
                  knots = list(hour = c(0, 24), dow = c(0.5, 7.5))))
```

## Fit

```{r m2_fit}
m2 <- brm(
  f_m2,
  data    = train_df,
  family  = student(),
  prior   = base_priors,
  knots   = list(hour = c(0, 24), dow = c(0.5, 7.5)),
  iter    = 2000, warmup = 1000, chains = 4, seed = 42,
  control = list(adapt_delta = 0.95),
  file    = "models/brms_m2"
)
```

```{r m2_ppc, fig.height=3}
pp_check(m2, ndraws = 50, type = "dens_overlay") +
  coord_cartesian(xlim = c(-600, 600)) +
  labs(title = "M2 PPC")
```

---

# LOO comparison

```{r loo}
m0 <- add_criterion(m0, "loo")
m1 <- add_criterion(m1, "loo")
m2 <- add_criterion(m2, "loo")

loo_compare(m0, m1, m2)
```

---

# Temporal holdout evaluation

LOO is in-sample and assumes exchangeability. This evaluates on days
the model never saw вЂ” the honest test.

```{r temporal_eval}
# dow is pre-factored with levels 1:7 in the data, so all levels are always present вЂ”
# no factor dimension mismatch when brms rebuilds the design matrix for newdata.
# brms handles unseen trip_id/stop_id with allow_new_levels = TRUE
pred_m2 <- posterior_predict(m2, newdata = test_df,
                              allow_new_levels = TRUE, ndraws = 500)

# Point estimate: posterior mean of predictive distribution
pred_mean <- colMeans(pred_m2)

# 90 % predictive interval
ci_lower <- apply(pred_m2, 2, quantile, 0.05)
ci_upper <- apply(pred_m2, 2, quantile, 0.95)

results <- test_df %>%
  mutate(
    predicted = pred_mean,
    ci_lower  = ci_lower,
    ci_upper  = ci_upper,
    residual  = delay_seconds - predicted,
    covered   = delay_seconds >= ci_lower & delay_seconds <= ci_upper
  )

cat(sprintf("MAE:       %.1f seconds\n", mean(abs(results$residual))))
cat(sprintf("RMSE:      %.1f seconds\n", sqrt(mean(results$residual^2))))
cat(sprintf("90%% Coverage: %.1f%%\n", 100 * mean(results$covered)))
```

```{r temporal_plots, fig.height=3}
p_pred <- ggplot(results, aes(x = delay_seconds, y = predicted)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_abline(slope = 1, linetype = "dashed") +
  labs(title = "Temporal holdout: observed vs predicted (M2)",
       x = "Observed delay (s)", y = "Predicted delay (s)") +
  theme_minimal()

p_resid <- ggplot(results, aes(x = factor(hour), y = residual)) +
  geom_boxplot(fill = "lightblue", outlier.alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals by hour", x = "Hour", y = "Residual (s)") +
  theme_minimal()

p_pred + p_resid
```

---

# Model validation

```{r model_checks, cache=FALSE}
cat("=== Model Validation: M2 ===\n\n")

# в”Ђв”Ђ Convergence в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
smry     <- summary(m2)
all_rhat <- c(smry$fixed[, "Rhat"], smry$spec_pars[, "Rhat"])
all_ess  <- c(smry$fixed[, "Bulk_ESS"], smry$fixed[, "Tail_ESS"],
              smry$spec_pars[, "Bulk_ESS"], smry$spec_pars[, "Tail_ESS"])
rhat_max <- max(all_rhat, na.rm = TRUE)
ess_min  <- min(all_ess,  na.rm = TRUE)

np        <- bayesplot::nuts_params(m2)
n_divs    <- sum(np$Value[np$Parameter == "divergent__"])
post_draws <- (m2$fit@sim$iter - m2$fit@sim$warmup) * m2$fit@sim$chains
div_pct   <- 100 * n_divs / post_draws

cat(sprintf("Rhat max:          %.4f  %s\n",
            rhat_max, ifelse(rhat_max > 1.05, "[WARN > 1.05]", "[OK]")))
cat(sprintf("ESS min:           %d    %s\n",
            as.integer(ess_min), ifelse(ess_min < 400, "[WARN < 400]", "[OK]")))
cat(sprintf("Divergent draws:   %d / %d  (%.1f%%)  %s\n",
            n_divs, post_draws, div_pct,
            ifelse(div_pct > 2, "[WARN > 2%]", "[OK]")))

# в”Ђв”Ђ Posterior sanity checks в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
fe           <- fixef(m2)
b_prev_mean  <- fe["previous_stop_delay", "Estimate"]
b_prev_q2.5  <- fe["previous_stop_delay", "Q2.5"]
b_prev_q97.5 <- fe["previous_stop_delay", "Q97.5"]
sigma_est    <- smry$spec_pars["sigma", "Estimate"]
nu_est       <- smry$spec_pars["nu",    "Estimate"]
nu_l95       <- smry$spec_pars["nu",    "l-95% CI"]

cat(sprintf("\nb_prev_delay:      %.3f [%.3f, %.3f]  %s\n",
            b_prev_mean, b_prev_q2.5, b_prev_q97.5,
            ifelse(b_prev_mean < 0.8 | b_prev_mean > 1.2,
                   "[WARN: outside [0.8, 1.2] вЂ” check data]", "[OK]")))
cat(sprintf("sigma:             %.1fs   %s\n",
            sigma_est,
            ifelse(sigma_est < 5 | sigma_est > 200, "[WARN: outside [5, 200]s]", "[OK]")))
cat(sprintf("nu:                %.2f  (l-95%% CI=%.2f)  %s\n",
            nu_est, nu_l95,
            ifelse(nu_l95 < 2.1, "[WARN: nu at lower bound вЂ” extremely heavy-tailed]",
                   "[OK]")))

# в”Ђв”Ђ Holdout checks в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
mae         <- mean(abs(results$residual))
coverage_90 <- mean(results$covered)
obs_mean    <- mean(test_df$delay_seconds)
pred_mean   <- mean(results$predicted)

cat(sprintf("\nMAE:               %.1fs  %s\n",
            mae, ifelse(mae > 120, "[WARN > 120s]", "[OK]")))
cat(sprintf("Coverage (90%%):    %.1f%%  %s\n",
            100 * coverage_90,
            ifelse(coverage_90 < 0.75 | coverage_90 > 0.95,
                   "[WARN: outside [75%, 95%]]", "[OK]")))
cat(sprintf("Obs mean: %.1fs  |  Pred mean: %.1fs  |  Gap: %.1fs  %s\n",
            obs_mean, pred_mean, abs(obs_mean - pred_mean),
            ifelse(abs(obs_mean - pred_mean) > 5, "[WARN > 5s gap]", "[OK]")))

# в”Ђв”Ђ Verdict в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
n_flags <- sum(c(
  rhat_max > 1.05, ess_min < 400, div_pct > 2,
  b_prev_mean < 0.8 | b_prev_mean > 1.2,
  sigma_est < 5 | sigma_est > 200,
  nu_l95 < 2.1,
  mae > 120,
  coverage_90 < 0.75 | coverage_90 > 0.95
))
verdict <- if (n_flags == 0) "PASS" else if (n_flags <= 2) "WARN" else "FAIL"
cat(sprintf("\n>>> Verdict: %s  (%d flag(s))\n", verdict, n_flags))
```

```{r run_tracker, cache=FALSE}
source("run_tracker.R")
track_run(
  model      = m2,
  train_df   = train_df,
  test_df    = test_df,
  results    = results,
  model_file = "models/brms_m2.rds",
  n_routes   = NA_integer_
)
```

---

# Prior predictive check (M2)

Fit M2 with `sample_prior = "only"` to see what delays the model
*expects before seeing any data*. If the prior implies impossible delays
(e.g., buses arriving 10 hours early), priors need tightening.

```{r prior_pc}
m2_prior <- update(m2, sample_prior = "only",
                   file = "models/brms_m2_prior_only")

pp_check(m2_prior, ndraws = 50, type = "dens_overlay") +
  coord_cartesian(xlim = c(-2000, 2000)) +
  labs(title = "Prior predictive check: what does the model expect before seeing data?")
```

---

# Notes: Multi-route model (when v2 data accumulates)

Once several weeks of data are collected via the v2 pipeline, the next model
adds a route-level grouping above trips. Trips on the same route share a
common baseline; different routes can still differ.

```{r multiRoute_preview, eval=FALSE}
# Load multi-route parquet when ready:
# df_multi <- arrow::read_parquet("../exports/all_routes_YYYY-MM-DD.parquet")

f_m3 <- bf(
  delay_seconds ~ previous_stop_delay + shape_dist_traveled +
    s(hour, bs = "cc", k = 8) +
    s(dow, bs = "cc", k = 5) +
    (1 | route_id / trip_id) +   # nested: trips within routes
    (1 | stop_id)
)

# Always inspect Stan code first
cat(make_stancode(f_m3, data = df_multi, family = student(),
                  prior = base_priors,
                  knots = list(hour = c(0, 24), dow = c(0.5, 7.5))))

# m3 <- brm(f_m3, data = df_multi, family = student(), prior = base_priors,
#           knots = list(hour = c(0, 24), dow = c(0.5, 7.5)),
#           iter = 2000, chains = 4, seed = 42,
#           control = list(adapt_delta = 0.95),
#           file = "models/brms_m3_multiRoute")
```
