---
title: "TransLink Multi-Route Delay Analysis — M3"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = TRUE, cache.lazy = FALSE)
```

# Setup

```{r libs, cache=FALSE}
library(brms)
library(dplyr)
library(ggplot2)
library(lubridate)
library(bayesplot)
library(loo)
library(arrow)
library(tidyr)
library(patchwork)

options(mc.cores = parallel::detectCores())
```

# Data snapshot

```{r load}
pq_path <- tail(
  sort(list.files("../exports", pattern = "all_routes_.*\\.parquet",
                  full.names = TRUE)),
  1
)
if (length(pq_path) == 0) stop("No all_routes_*.parquet found in exports/")
cat(sprintf("Loading: %s\n", pq_path))

raw <- arrow::read_parquet(pq_path) |>
  filter(
    route_id != "",
    !is.na(delay_seconds),
    !is.na(previous_stop_delay),
    !is.na(shape_dist_traveled),
    !is.na(hour),
    !is.na(stop_id),
    !is.na(trip_id),
    abs(delay_seconds)        <= 3600,
    abs(previous_stop_delay)  <= 3600
  ) |>
  mutate(
    timestamp   = as.POSIXct(timestamp, tz = "UTC"),
    # `date` = Vancouver local date to match Python pipeline's Vancouver-tz
    # service_date (audit 2026-07-08). Bare as.Date() used session TZ silently.
    date        = as.Date(format(timestamp, "%Y-%m-%d", tz = "America/Vancouver")),
    hour        = as.integer(hour),
    dow         = as.integer(wday(timestamp, week_start = 1L)),
    trip_id     = as.character(trip_id),
    stop_id     = as.character(stop_id),
    route_id    = as.character(route_id)
  )

cat("=== Data snapshot ===\n")
cat(sprintf("Date range:   %s → %s\n", min(raw$date), max(raw$date)))
cat(sprintf("Total rows:   %s\n", format(nrow(raw), big.mark=",")))
cat(sprintf("Routes:       %d\n", n_distinct(raw$route_id)))
cat(sprintf("Trips:        %s\n", format(n_distinct(raw$trip_id), big.mark=",")))
cat(sprintf("Stops:        %s\n", format(n_distinct(raw$stop_id), big.mark=",")))
cat(sprintf("Mean delay:   %.1f s\n", mean(raw$delay_seconds, na.rm=TRUE)))
```

```{r data_quality, cache=FALSE}
cat("=== Data Quality ===\n\n")

n_before_anomaly <- nrow(raw)

# NULL check — should all be 0 after filtering
model_vars <- c("delay_seconds", "previous_stop_delay", "shape_dist_traveled",
                "hour", "dow", "trip_id", "stop_id", "route_id")
null_rates <- sapply(model_vars, function(v) mean(is.na(raw[[v]])))
if (any(null_rates > 0)) {
  cat("WARN: Non-zero NULL rates after filtering:\n")
  print(null_rates[null_rates > 0])
} else {
  cat("NULL check: all model variables clean (0% NULL)\n")
}

# Data-driven anomaly day detection (replaces hardcoded date exclusions)
daily_stats <- raw %>%
  group_by(date) %>%
  summarise(n = n(), mean_delay = mean(delay_seconds, na.rm = TRUE), .groups = "drop")

overall_mean <- mean(daily_stats$mean_delay)
overall_sd   <- sd(daily_stats$mean_delay)

anomaly_days <- daily_stats %>%
  filter(n < 0.20 * median(daily_stats$n) |
         abs(mean_delay - overall_mean) > 2 * overall_sd) %>%
  pull(date)

cat(sprintf("Anomaly days excluded: %s\n",
            if (length(anomaly_days) > 0) paste(anomaly_days, collapse = ", ") else "none"))

if (length(anomaly_days) > 0) {
  raw <- raw %>% filter(!date %in% anomaly_days)
  cat(sprintf("Rows: %s → %s  (lost %s)\n",
              format(n_before_anomaly, big.mark = ","),
              format(nrow(raw), big.mark = ","),
              format(n_before_anomaly - nrow(raw), big.mark = ",")))
}

# Delay distribution summary
delay_mean <- mean(raw$delay_seconds)
delay_sd   <- sd(raw$delay_seconds)
skew       <- mean(((raw$delay_seconds - delay_mean) / delay_sd)^3)
cat(sprintf("\nDelay distribution: mean=%.1fs  SD=%.1fs  skewness=%.2f\n",
            delay_mean, delay_sd, skew))
cat(sprintf("Date range: %s → %s  (%d unique dates)\n",
            min(raw$date), max(raw$date), n_distinct(raw$date)))
```

## Filter to routes with sufficient data

M3 needs enough observations per route to estimate route-level random effects.
Require ≥ 500 rows; subsample 500 rows per route across the top 25 routes.

```{r anomaly_routes, cache=FALSE}
# Curated exclusion list.  Two distinct failure modes — do not conflate:
#
# (A) Stop-sequence tie artifacts: RT feed broadcasts multiple stops at the
#     same stop_sequence number (branching/loop topology).  LAG()-based
#     previous_stop_delay is incorrect at tie points.  Do NOT replace with a
#     threshold-based auto-exclusion — the bad-trip distribution is continuous
#     (0–98%), not bimodal.  Update when the diagnostic below flags a new
#     route above ~60% with >= 50 trips.
#     Identified via stop-sequence integrity analysis (2026-05-29):
#       12940  98.5%   6700  88.1%   30055  63.6%   6702  60.6%
#     6718 stays excluded despite appearing at 27.3% in recent processed_stops:
#       only 11 trips in May-16+ window — historically 100% bad.  Coverage
#       artifact.
#
# (B) Structural GTFS join failure (different root cause from A):
#     6619 (Route 009 Broadway/Lougheed): 65.5% of stops lack
#     shape_dist_traveled because whole route sections — the Lougheed Hwy
#     extension eastbound (stops 901–908) and both termini — have no shape
#     entry in gtfs_static.db.  The NULL pattern is geographic, not random.
#     After pre-filtering on shape_dist, only the central Broadway corridor
#     remains — a structurally biased, non-representative sample.  Excluded
#     here so a future data-volume increase cannot silently re-admit it.
anomaly_routes <- c("12940", "6700", "30055", "6702", "6718", "6619")

# Holdout routes — excluded from training, used only for out-of-distribution
# evaluation after fitting.  Chosen for benchmark value and geographic contrast:
#   6641 (099 Broadway B-Line): the route used for all single-route hand-tuning
#     (M0–M3 in brms_analysis.Rmd) — allows direct Ralph loop vs hand-built
#     model comparison.
#   6705 (321 King George Blvd): suburban Surrey route — geographic contrast
#     to the urban B-Line tests generalization across route types.
holdout_routes <- c("6641", "6705")

# Diagnostic: print the full tied-sequence rate table so new candidates are
# visible on every render.  Routes above ~60% with solid trip counts warrant
# review for addition to anomaly_routes.
report_seq_integrity <- function(df, min_trips = 10) {
  trip_flags <- df |>
    group_by(.data$route_id, .data$trip_id) |>
    summarise(
      has_tie = any(duplicated(.data$stop_sequence)),
      .groups = "drop"
    )

  route_flags <- trip_flags |>
    group_by(.data$route_id) |>
    summarise(
      total_trips = n(),
      bad_trips   = sum(.data$has_tie),
      pct_bad     = round(100 * mean(.data$has_tie), 1),
      .groups     = "drop"
    ) |>
    filter(.data$total_trips >= min_trips) |>
    arrange(desc(.data$pct_bad))

  cat("Stop-sequence tie rates by route (routes with >= 10 trips):\n")
  cat(sprintf("  %-10s  %10s  %9s  %7s  %s\n",
              "route_id", "total_trips", "bad_trips", "pct_bad", "excluded"))
  for (i in seq_len(nrow(route_flags))) {
    r <- route_flags[i, ]
    excl <- if (r$route_id %in% anomaly_routes) "<-- EXCLUDED" else ""
    cat(sprintf("  %-10s  %10d  %9d  %6.1f%%  %s\n",
                r$route_id, r$total_trips, r$bad_trips, r$pct_bad, excl))
  }

  invisible(route_flags)
}

report_seq_integrity(raw)
cat(sprintf("\nExcluding %d route(s): %s\n",
            length(anomaly_routes), paste(anomaly_routes, collapse = ", ")))
```

```{r subsample}
set.seed(42)

holdout_df <- raw |>
  filter(route_id %in% holdout_routes)
cat(sprintf(
  "Holdout set: %s rows across %d routes (%s)\n",
  format(nrow(holdout_df), big.mark = ","),
  n_distinct(holdout_df$route_id),
  paste(sort(unique(holdout_df$route_id)), collapse = ", ")
))

route_counts <- raw |>
  filter(!route_id %in% anomaly_routes, !route_id %in% holdout_routes) |>
  count(route_id, name = "n_rows") |>
  filter(n_rows >= 500) |>
  arrange(desc(n_rows))

cat(sprintf("Routes with ≥500 rows (excl. %d anomaly routes): %d\n",
            length(anomaly_routes), nrow(route_counts)))

# Log routes on the edge of the 500-row threshold (±100 rows)
edge_routes <- raw |>
  filter(!route_id %in% anomaly_routes, !route_id %in% holdout_routes) |>
  count(route_id) |>
  filter(n >= 400 & n < 600) |>
  arrange(n)
if (nrow(edge_routes) > 0) {
  cat(sprintf("Routes near ≥500 threshold (400–599 rows): %s\n",
              paste(sprintf("%s(%d)", edge_routes$route_id, edge_routes$n), collapse = ", ")))
}

# Top 25 routes × 500 rows = ~12,500 rows.
# Gives ~15 trips per route to estimate route-level random effects vs ~4 at 150 rows.
top_routes <- route_counts |> slice_max(n_rows, n = 25)

df <- raw |>
  filter(route_id %in% top_routes$route_id) |>
  group_by(route_id) |>
  slice_sample(n = 500) |>
  ungroup()

cat(sprintf("After 500-row-per-route subsample (top 25): %s rows, %d routes\n",
            format(nrow(df), big.mark = ","), n_distinct(df$route_id)))
```

## Temporal split

```{r split, cache=FALSE}
dates    <- sort(unique(df$date))
cutoff   <- dates[floor(0.75 * length(dates))]
train_df <- filter(df, date <= cutoff)
test_df  <- filter(df, date >  cutoff)

cat(sprintf("Train: %s rows up to %s\n", format(nrow(train_df), big.mark=","), cutoff))
cat(sprintf("Test:  %s rows from %s → %s\n",
            format(nrow(test_df), big.mark=","), min(test_df$date), max(test_df$date)))
```

---

# Priors

```{r priors}
base_priors <- c(
  prior(normal(0, 2),       class = b),
  prior(normal(0, 10),      class = Intercept),
  prior(exponential(1),     class = sigma),
  prior(exponential(1),     class = sd),
  prior(gamma(2, 0.1),      class = nu, lb = 2)
)
```

---

# M3 — Multi-route nested model

Route-level random intercepts: `(1 | route_id)`.
Routes share the same `previous_stop_delay` slope — the key question is whether
the intercepts vary systematically across routes.

## Inspect Stan code

```{r m3_stancode}
f_m3 <- bf(
  delay_seconds ~ previous_stop_delay + shape_dist_traveled +
    s(hour, bs = "cc", k = 8) +
    s(dow,  bs = "cc", k = 5) +
    (1 | route_id) +
    (1 | trip_id) +
    (1 | stop_id)
)

cat(make_stancode(f_m3, data = train_df, family = student(),
                  prior = base_priors,
                  knots = list(hour = c(0, 24), dow = c(0.5, 7.5))))
```

## Fit

```{r m3_fit}
m3 <- brm(
  f_m3,
  data    = train_df,
  family  = student(),
  prior   = base_priors,
  knots   = list(hour = c(0, 24), dow = c(0.5, 7.5)),
  iter    = 2000, warmup = 1000, chains = 4, seed = 42,
  control = list(adapt_delta = 0.95),
  file    = "models/brms_m3_multiRoute",
  refresh = 50
)
```

## Summary

```{r m3_summary}
summary(m3, pars = c("b_Intercept", "b_previous_stop_delay",
                      "b_shape_dist_traveled", "sigma", "nu"))
```

## Diagnostics

```{r m3_diag, fig.height=3}
mcmc_trace(as.array(m3), pars = c("b_Intercept", "b_previous_stop_delay",
                                    "b_shape_dist_traveled", "sigma", "nu"))
```

```{r m3_ppc, fig.height=3}
pp_check(m3, ndraws = 50, type = "dens_overlay") +
  coord_cartesian(xlim = c(-600, 600)) +
  labs(title = "M3 PPC: 50 posterior predictive draws vs observed")
```

---

# Route-level random effects

Which routes are systematically delayed or ahead of schedule,
after accounting for delay propagation, distance, hour, and day of week?

```{r route_effects, fig.height=10}
re <- ranef(m3)$route_id %>%
  as.data.frame() %>%
  tibble::rownames_to_column("route_id") %>%
  rename(Estimate = Estimate.Intercept,
         Q2.5     = Q2.5.Intercept,
         Q97.5    = Q97.5.Intercept) %>%
  arrange(Estimate)

# Attach row counts
re <- re %>%
  left_join(route_counts %>% select(route_id, n_rows), by = "route_id")

ggplot(re, aes(x = Estimate, y = reorder(route_id, Estimate))) +
  geom_point(aes(size = n_rows), color = "steelblue") +
  geom_errorbarh(aes(xmin = Q2.5, xmax = Q97.5), height = 0.3, alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_size_continuous(name = "obs in period", range = c(1, 4)) +
  labs(title = "Route-level random intercepts (M3)",
       subtitle = "Positive = systematically later than average; Negative = systematically earlier",
       x = "Effect (seconds)", y = "Route ID") +
  theme_minimal(base_size = 9)
```

## Distribution of route effects

```{r route_dist, fig.height=3}
ggplot(re, aes(x = Estimate)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(title = "Distribution of route-level intercepts",
       x = "Effect (seconds)", y = "Count") +
  theme_minimal()
```

---

# Temporal holdout

```{r m3_holdout}
pred_m3 <- posterior_predict(m3, newdata = test_df,
                              allow_new_levels = TRUE, ndraws = 500)

pred_mean <- colMeans(pred_m3)
ci_lower  <- apply(pred_m3, 2, quantile, 0.05)
ci_upper  <- apply(pred_m3, 2, quantile, 0.95)

results <- test_df %>%
  mutate(
    predicted = pred_mean,
    ci_lower  = ci_lower,
    ci_upper  = ci_upper,
    residual  = delay_seconds - predicted,
    covered   = delay_seconds >= ci_lower & delay_seconds <= ci_upper
  )

cat(sprintf("MAE:          %.1f seconds\n", mean(abs(results$residual))))
cat(sprintf("RMSE:         %.1f seconds\n", sqrt(mean(results$residual^2))))
cat(sprintf("90%% Coverage: %.1f%%\n",  100 * mean(results$covered)))
```

```{r m3_holdout_plots, fig.height=3}
p_pred <- ggplot(results, aes(x = delay_seconds, y = predicted)) +
  geom_point(alpha = 0.1, color = "steelblue") +
  geom_abline(slope = 1, linetype = "dashed") +
  labs(title = "Temporal holdout: observed vs predicted (M3)",
       x = "Observed delay (s)", y = "Predicted delay (s)") +
  theme_minimal()

p_resid <- ggplot(results, aes(x = factor(hour), y = residual)) +
  geom_boxplot(fill = "lightblue", outlier.alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals by hour", x = "Hour", y = "Residual (s)") +
  theme_minimal()

p_pred + p_resid
```

---

# Route holdout evaluation

Routes 6641 (099 Broadway B-Line) and 6705 (321 King George Blvd) were never
seen during training. This is the primary generalization test — not a temporal
slice of training routes, but entirely unseen routes. Route 6641 also allows
direct comparison against the hand-built M0–M3 single-route models.

```{r m3_holdout_routes}
pred_holdout <- posterior_predict(
  m3, newdata = holdout_df,
  allow_new_levels = TRUE, ndraws = 500
)

holdout_results <- holdout_df |>
  mutate(
    predicted = colMeans(pred_holdout),
    ci_lower  = apply(pred_holdout, 2, quantile, 0.05),
    ci_upper  = apply(pred_holdout, 2, quantile, 0.95),
    residual  = delay_seconds - predicted,
    covered   = delay_seconds >= ci_lower & delay_seconds <= ci_upper
  )

for (r in sort(unique(holdout_results$route_id))) {
  rd <- filter(holdout_results, route_id == r)
  cat(sprintf(
    "Route %-6s  MAE=%.1fs  RMSE=%.1fs  90%%CI=%.1f%%  rows=%d\n",
    r,
    mean(abs(rd$residual)),
    sqrt(mean(rd$residual^2)),
    100 * mean(rd$covered),
    nrow(rd)
  ))
}

cat(sprintf(
  "\nOverall  MAE=%.1fs  RMSE=%.1fs  90%%CI=%.1f%%\n",
  mean(abs(holdout_results$residual)),
  sqrt(mean(holdout_results$residual^2)),
  100 * mean(holdout_results$covered)
))
```

```{r m3_holdout_routes_plot, fig.height=3}
ggplot(holdout_results, aes(x = delay_seconds, y = predicted,
                             colour = route_id)) +
  geom_point(alpha = 0.15) +
  geom_abline(slope = 1, linetype = "dashed") +
  scale_colour_brewer(palette = "Set1", name = "Route") +
  labs(title = "Route holdout: observed vs predicted (M3)",
       subtitle = "Routes 6641 and 6705 — never seen during training",
       x = "Observed delay (s)", y = "Predicted delay (s)") +
  theme_minimal()
```

---

# Model validation

```{r model_checks, cache=FALSE}
cat("=== Model Validation: M3 ===\n\n")

# ── Convergence ──────────────────────────────────────────────────────────────
smry      <- summary(m3)
re_all    <- do.call(rbind, smry$random)   # combines route_id, trip_id, stop_id
all_rhat  <- c(smry$fixed[, "Rhat"], smry$spec_pars[, "Rhat"],
               re_all[, "Rhat"])
all_ess   <- c(smry$fixed[, "Bulk_ESS"], smry$fixed[, "Tail_ESS"],
               smry$spec_pars[, "Bulk_ESS"], smry$spec_pars[, "Tail_ESS"],
               re_all[, "Bulk_ESS"], re_all[, "Tail_ESS"])
rhat_max  <- max(all_rhat, na.rm = TRUE)
ess_min   <- min(all_ess,  na.rm = TRUE)

np         <- bayesplot::nuts_params(m3)
n_divs     <- sum(np$Value[np$Parameter == "divergent__"])
post_draws <- (m3$fit@sim$iter - m3$fit@sim$warmup) * m3$fit@sim$chains
div_pct    <- 100 * n_divs / post_draws

cat(sprintf("Rhat max:          %.4f  %s\n",
            rhat_max, ifelse(rhat_max > 1.05, "[WARN > 1.05]", "[OK]")))
cat(sprintf("ESS min:           %d    %s\n",
            as.integer(ess_min), ifelse(ess_min < 400, "[WARN < 400]", "[OK]")))
cat(sprintf("Divergent draws:   %d / %d  (%.1f%%)  %s\n",
            n_divs, post_draws, div_pct,
            ifelse(div_pct > 2, "[WARN > 2%]", "[OK]")))

# ── Posterior sanity checks ──────────────────────────────────────────────────
fe           <- fixef(m3)
b_prev_mean  <- fe["previous_stop_delay", "Estimate"]
b_prev_q2.5  <- fe["previous_stop_delay", "Q2.5"]
b_prev_q97.5 <- fe["previous_stop_delay", "Q97.5"]
sigma_est    <- smry$spec_pars["sigma", "Estimate"]
nu_est       <- smry$spec_pars["nu",    "Estimate"]
nu_l95       <- smry$spec_pars["nu",    "l-95% CI"]
route_sd     <- smry$random$route_id["sd(Intercept)", "Estimate"]
route_sd_l95 <- smry$random$route_id["sd(Intercept)", "l-95% CI"]
trip_sd      <- smry$random$trip_id["sd(Intercept)",  "Estimate"]
stop_sd      <- smry$random$stop_id["sd(Intercept)",  "Estimate"]

cat(sprintf("\nb_prev_delay:      %.3f [%.3f, %.3f]  %s\n",
            b_prev_mean, b_prev_q2.5, b_prev_q97.5,
            ifelse(b_prev_mean < 0.8 | b_prev_mean > 1.2,
                   "[WARN: outside [0.8, 1.2] — check data]", "[OK]")))
cat(sprintf("sigma:             %.1fs   %s\n",
            sigma_est,
            ifelse(sigma_est < 5 | sigma_est > 200, "[WARN: outside [5, 200]s]", "[OK]")))
cat(sprintf("nu:                %.2f  (l-95%% CI=%.2f)  %s\n",
            nu_est, nu_l95,
            ifelse(nu_l95 < 2.1, "[WARN: nu at lower bound — extremely heavy-tailed]",
                   "[OK]")))
cat(sprintf("Route SD:          %.1fs  (l-95%% CI=%.1fs)  %s\n",
            route_sd, route_sd_l95,
            ifelse(route_sd_l95 < 0.5,
                   "[WARN: route SD near zero — no detectable route variation]", "[OK]")))
cat(sprintf("Trip SD:           %.1fs\n", trip_sd))
cat(sprintf("Stop SD:           %.1fs\n", stop_sd))

# ── Holdout checks ───────────────────────────────────────────────────────────
mae         <- mean(abs(results$residual))
coverage_90 <- mean(results$covered)
obs_mean    <- mean(test_df$delay_seconds)
pred_mean   <- mean(results$predicted)

cat(sprintf("\nMAE:               %.1fs  %s\n",
            mae, ifelse(mae > 150, "[WARN > 150s (multi-route threshold)]", "[OK]")))
cat(sprintf("Coverage (90%%):    %.1f%%  %s\n",
            100 * coverage_90,
            ifelse(coverage_90 < 0.75 | coverage_90 > 0.95,
                   "[WARN: outside [75%, 95%]]", "[OK]")))
cat(sprintf("Obs mean: %.1fs  |  Pred mean: %.1fs  |  Gap: %.1fs  %s\n",
            obs_mean, pred_mean, abs(obs_mean - pred_mean),
            ifelse(abs(obs_mean - pred_mean) > 5, "[WARN > 5s gap]", "[OK]")))

# ── Verdict ──────────────────────────────────────────────────────────────────
n_flags <- sum(c(
  rhat_max > 1.05, ess_min < 400, div_pct > 2,
  b_prev_mean < 0.8 | b_prev_mean > 1.2,
  sigma_est < 5 | sigma_est > 200,
  nu_l95 < 2.1,
  route_sd_l95 < 0.5,
  mae > 150,
  coverage_90 < 0.75 | coverage_90 > 0.95
))
# trip_sd and stop_sd are informational — not flagged, but tracked in run_tracker
verdict <- if (n_flags == 0) "PASS" else if (n_flags <= 2) "WARN" else "FAIL"
cat(sprintf("\n>>> Verdict: %s  (%d flag(s))\n", verdict, n_flags))
```

```{r run_tracker, cache=FALSE}
source("run_tracker.R")
track_run(
  model      = m3,
  train_df   = train_df,
  test_df    = test_df,
  results    = results,
  model_file = "models/brms_m3_multiRoute.rds",
  n_routes   = n_distinct(train_df$route_id)
)
```
