---
title: "TransLink Bayesian Delay — Visualization Showcase"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = TRUE, cache.lazy = FALSE,
                      fig.width = 10, fig.height = 5)
```

```{r libs, cache=FALSE}
library(brms)
library(dplyr)
library(ggplot2)
library(lubridate)
library(bayesplot)
library(patchwork)
library(arrow)
library(DBI)
library(RSQLite)
library(scales)

options(mc.cores = parallel::detectCores())
theme_set(theme_minimal(base_size = 13))
```

```{r load_models, cache=TRUE}
m0 <- readRDS("models/brms_m0.rds")
m1 <- readRDS("models/brms_m1.rds")
m2 <- readRDS("models/brms_m2.rds")
m3 <- readRDS("models/brms_m3_multiRoute.rds")
```

```{r load_data, cache=TRUE}
# Auto-discover latest parquets so this file doesn't need manual date updates
pq_6641 <- tail(sort(list.files("../exports", pattern = "route6641_dir0_.*\\.parquet",
                                 full.names = TRUE)), 1)
pq_all  <- tail(sort(list.files("../exports", pattern = "all_routes_.*\\.parquet",
                                 full.names = TRUE)), 1)

filter_and_clean <- function(df) {
  df %>%
    filter(
      !is.na(delay_seconds), !is.na(previous_stop_delay),
      !is.na(shape_dist_traveled), !is.na(hour),
      !is.na(stop_id), !is.na(trip_id),
      abs(delay_seconds)       <= 3600,
      abs(previous_stop_delay) <= 3600
    ) %>%
    mutate(
      timestamp = as.POSIXct(timestamp, tz = "UTC"),
      date      = as.Date(timestamp),
      hour      = as.integer(hour),
      dow       = as.integer(wday(timestamp, week_start = 1L)),
      trip_id   = as.character(trip_id),
      stop_id   = as.character(stop_id)
    )
}

anomaly_exclude <- function(df) {
  daily <- df %>% group_by(date) %>%
    summarise(n = n(), mean_d = mean(delay_seconds, na.rm = TRUE), .groups = "drop")
  gm <- mean(daily$mean_d)
  gs <- sd(daily$mean_d)
  bad <- daily %>%
    filter(n < 0.20 * median(daily$n) | abs(mean_d - gm) > 2 * gs) %>%
    pull(date)
  df %>% filter(!date %in% bad)
}

raw6641 <- read_parquet(pq_6641) %>% filter_and_clean() %>% anomaly_exclude()
rawAll  <- read_parquet(pq_all)  %>%
  filter(route_id != "") %>%
  filter_and_clean() %>%
  anomaly_exclude() %>%
  mutate(route_id = as.character(route_id))

# Train / test split for route 6641
dates6641  <- sort(unique(raw6641$date))
cutoff6641 <- dates6641[floor(0.75 * length(dates6641))]
train6641  <- filter(raw6641, date <= cutoff6641)
test6641   <- filter(raw6641, date >  cutoff6641)

cat(sprintf("Route 6641 — train: %d rows  test: %d rows\n",
            nrow(train6641), nrow(test6641)))
cat(sprintf("All routes  — %s rows, %d routes\n",
            format(nrow(rawAll), big.mark = ","),
            n_distinct(rawAll$route_id)))
```

---

# Viz 1 — Delay propagates 1:1 across stops

**What:** Scatter of current stop delay vs previous stop delay, with the posterior
mean slope and 90 % credible band.
**Why:** The slope ≈ 1.0 is the headline finding — one second late at stop N
predicts one second late at stop N+1, across all hours and routes.

```{r viz1_grid, cache=TRUE}
psd_range <- quantile(train6641$previous_stop_delay, c(0.02, 0.98))
pred_grid <- data.frame(
  previous_stop_delay = seq(psd_range[1], psd_range[2], length.out = 120),
  shape_dist_traveled = median(train6641$shape_dist_traveled),
  hour                = 8L,   # fix at 8am
  dow                 = 2L,   # fix at Tuesday
  trip_id             = train6641$trip_id[1],
  stop_id             = train6641$stop_id[1]
)

fv <- fitted(m2, newdata = pred_grid, re_formula = NA,
             allow_new_levels = TRUE, probs = c(0.05, 0.95))
pred_grid <- pred_grid %>%
  mutate(fit = fv[, "Estimate"], lo = fv[, "Q5"], hi = fv[, "Q95"])
```

```{r viz1_plot, fig.height=5, cache=FALSE}
set.seed(1)
scatter_sample <- slice_sample(test6641, n = min(3000, nrow(test6641)))

b     <- fixef(m2)["previous_stop_delay", "Estimate"]
b_lo  <- fixef(m2)["previous_stop_delay", "Q2.5"]
b_hi  <- fixef(m2)["previous_stop_delay", "Q97.5"]

ggplot() +
  geom_point(data = scatter_sample,
             aes(x = previous_stop_delay, y = delay_seconds),
             alpha = 0.12, size = 0.9, color = "steelblue") +
  geom_ribbon(data = pred_grid,
              aes(x = previous_stop_delay, ymin = lo, ymax = hi),
              fill = "coral", alpha = 0.35) +
  geom_line(data = pred_grid,
            aes(x = previous_stop_delay, y = fit),
            color = "coral", linewidth = 1.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", alpha = 0.4) +
  coord_cartesian(xlim = c(-250, 250), ylim = c(-250, 250)) +
  labs(
    title   = "Delay propagates almost perfectly from stop to stop",
    subtitle = sprintf("Posterior slope = %.4f  [%.4f, %.4f]  (dashed = perfect 1:1)",
                       b, b_lo, b_hi),
    x = "Previous stop delay (s)",
    y = "Current stop delay (s)",
    caption = "Route 6641 — holdout test set  |  90% credible band at median hour/DOW"
  )
```

---

# Viz 2 — Hour-of-day effect: factor vs cyclic spline

**What:** How the model's prediction shifts by hour, comparing 24 raw dummy
variables (M0) against a smooth cyclic cubic spline (M1).
**Why:** M1 is better because midnight is not discontinuous — hour 23 should
flow back into hour 0. The spline enforces that.

```{r viz2_m1_grid, cache=TRUE}
hour_grid <- data.frame(
  hour                = 0:23,
  dow                 = 1L,
  previous_stop_delay = 0,
  shape_dist_traveled = median(train6641$shape_dist_traveled),
  trip_id             = train6641$trip_id[1],
  stop_id             = train6641$stop_id[1]
)
fv_m1 <- as.data.frame(fitted(m1, newdata = hour_grid, re_formula = NA,
                               allow_new_levels = TRUE))
hour_m1_df <- bind_cols(hour_grid, fv_m1)
```

```{r viz2_plot, cache=FALSE}
# M0: extract factor(hour) posterior means
hour_m0_df <- as.data.frame(fixef(m0)) %>%
  tibble::rownames_to_column("term") %>%
  filter(grepl("^factorhour", term)) %>%
  mutate(hour = as.integer(sub("factorhour", "", term))) %>%
  select(hour, Estimate, Q2.5, Q97.5) %>%
  bind_rows(data.frame(hour = 0L, Estimate = 0, Q2.5 = 0, Q97.5 = 0)) %>%
  arrange(hour)

p_m0 <- ggplot(hour_m0_df, aes(x = hour, y = Estimate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, fill = "steelblue") +
  geom_vline(xintercept = c(7, 9, 16, 18), linetype = "dotted", alpha = 0.4) +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(title = "M0: 24 independent dummies", x = "Hour", y = "Hour effect (s)")

p_m1 <- ggplot(hour_m1_df, aes(x = hour, y = Estimate)) +
  geom_line(color = "coral", linewidth = 1) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, fill = "coral") +
  geom_vline(xintercept = c(7, 9, 16, 18), linetype = "dotted", alpha = 0.4) +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(title = "M1: cyclic cubic spline (k = 8)",
       x = "Hour", y = "Hour effect (s)")

p_m0 + p_m1 +
  plot_annotation(
    title    = "Cyclic spline vs 24 dummies for time-of-day",
    subtitle = "Dotted lines = rush hour boundaries  |  Previous delay, distance fixed at median"
  )
```

---

# Viz 3 — Which routes run systematically late?

**What:** Posterior random intercepts from M3 for 50 routes, sorted by effect.
Positive = consistently later than average; negative = consistently earlier.
**Why:** After controlling for delay propagation, distance, and time patterns,
the residual route effect reveals structural differences (traffic, route length, stops).

```{r viz3_plot, fig.height=9}
re_df <- ranef(m3)$route_id %>%
  as.data.frame() %>%
  tibble::rownames_to_column("route_id") %>%
  rename(Estimate = Estimate.Intercept,
         Q2.5     = Q2.5.Intercept,
         Q97.5    = Q97.5.Intercept) %>%
  arrange(Estimate) %>%
  mutate(
    route_id = factor(route_id, levels = route_id),
    sign     = ifelse(Estimate > 0, "Late", "Early")
  )

ggplot(re_df, aes(x = Estimate, y = route_id, color = sign)) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = Q2.5, xmax = Q97.5), height = 0.4, alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.6) +
  scale_color_manual(values = c("Late" = "coral", "Early" = "steelblue"),
                     guide = "none") +
  labs(
    title    = "Route-level random intercepts (M3) — who runs late?",
    subtitle = "After controlling for delay propagation, distance, hour, and day of week",
    x        = "Systematic offset (seconds)",
    y        = "Route ID",
    caption  = "Points = posterior mean  |  Bars = 95% credible interval"
  )
```

---

# Viz 4 — Does the model actually fit the data?

**What:** 50 delay distributions simulated from the posterior (light lines) vs
the observed distribution (dark line). If the model fits, they should overlap.
**Why:** The standard Bayesian goodness-of-fit check. A Normal model would
fail here — the observed distribution has heavier tails.

```{r viz4_plot}
p_ppc_m2 <- pp_check(m2, ndraws = 50, type = "dens_overlay") +
  coord_cartesian(xlim = c(-400, 600)) +
  labs(title = "M2 — Route 6641") +
  theme(legend.position = "none")

p_ppc_m3 <- pp_check(m3, ndraws = 50, type = "dens_overlay") +
  coord_cartesian(xlim = c(-400, 600)) +
  labs(title = "M3 — 50 routes") +
  theme(legend.position = "bottom")

p_ppc_m2 + p_ppc_m3 +
  plot_annotation(
    title    = "Posterior predictive check: model vs observed delay distribution",
    subtitle = "Light lines = 50 draws from posterior predictive  |  Dark line = observed"
  )
```

---

# Viz 5 — Are findings stable week over week?

**What:** Key parameters (`b_previous_stop_delay`, MAE) plotted across run dates.
Each point is one weekly refresh. Error bars = 95% credible interval.
**Why:** The most important robustness check — if coefficients bounce wildly as data
accumulates, the model may be overfit to one week. Stability = credible findings.

```{r viz5_plot, cache=FALSE}
run_log <- read.csv("../exports/run_log.csv", stringsAsFactors = FALSE) %>%
  mutate(run_date  = as.Date(run_date),
         model_tag = ifelse(is.na(n_routes), "M2 (Route 6641)", "M3 (50 routes)"))

p_b <- ggplot(run_log, aes(x = run_date, y = b_prev_delay_mean,
                            ymin = b_prev_delay_q2.5, ymax = b_prev_delay_q97.5,
                            color = model_tag, fill = model_tag)) +
  geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.4) +
  geom_ribbon(alpha = 0.15, linewidth = 0) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  scale_color_manual(values = c("M2 (Route 6641)" = "steelblue",
                                "M3 (50 routes)"  = "coral")) +
  scale_fill_manual( values = c("M2 (Route 6641)" = "steelblue",
                                "M3 (50 routes)"  = "coral")) +
  labs(title = "Propagation coefficient over time",
       subtitle = "Should hover near 1.0 — drift indicates data shift",
       x = NULL, y = "b_prev_delay", color = NULL, fill = NULL) +
  theme(legend.position = "top")

p_mae <- ggplot(run_log, aes(x = run_date, y = mae,
                              color = model_tag, group = model_tag)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  scale_color_manual(values = c("M2 (Route 6641)" = "steelblue",
                                "M3 (50 routes)"  = "coral"),
                     guide = "none") +
  labs(title = "Holdout MAE over time",
       subtitle = "Lower = better; stability = model isn't overfitting to one period",
       x = "Run date", y = "MAE (seconds)")

p_b / p_mae +
  plot_annotation(
    caption = sprintf("Showing %d run(s) — add weekly refreshes to see trend",
                      nrow(run_log))
  )
```

---

# Viz 6 — When are buses late? Hour × Day heatmap

**What:** Mean delay for every combination of hour (x) and day of week (y)
across all 230 routes. Color = running late (red) vs early (blue).
**Why:** Makes the "cyclic time" structure concrete — rush hours stand out
visually, and the weekend pattern differs from weekdays.

```{r viz6_static}
heat_df <- rawAll %>%
  group_by(hour, dow) %>%
  summarise(mean_delay = mean(delay_seconds, na.rm = TRUE),
            n = n(), .groups = "drop") %>%
  mutate(dow_label = factor(dow, levels = 1:7,
                            labels = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")))

lim <- max(abs(heat_df$mean_delay), na.rm = TRUE)
lim <- min(lim, 120)  # cap color scale at ±2 min for readability

ggplot(heat_df, aes(x = hour, y = dow_label, fill = mean_delay)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = ifelse(abs(mean_delay) > 10,
                               sprintf("%+.0f", mean_delay), "")),
            size = 2.8, color = "white", fontface = "bold") +
  scale_fill_gradient2(
    low = "steelblue", mid = "white", high = "coral",
    midpoint = 0, limits = c(-lim, lim),
    oob = squish,
    name = "Mean delay (s)",
    labels = function(x) ifelse(x > 0, sprintf("+%ds", x), sprintf("%ds", x))
  ) +
  scale_x_continuous(breaks = seq(0, 23, 3),
                     labels = function(h) sprintf("%02d:00", h)) +
  labs(
    title    = "Mean delay by hour and day of week — all 230 routes",
    subtitle = "Red = running late  |  Blue = running early  |  White = on time",
    x = "Hour of day", y = NULL
  ) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r viz6_animated, eval=requireNamespace("gganimate", quietly=TRUE), fig.height=4}
# library(gganimate)

# # Animate delay density shifting across hours of day
# set.seed(42)
# anim_df <- rawAll %>%
#   filter(!is.na(hour), !is.na(delay_seconds)) %>%
#   group_by(hour) %>%
#   slice_sample(n = 800L) %>%
#   ungroup() %>%
#   mutate(hour_label = sprintf("Hour %02d:00", hour))

# anim <- ggplot(anim_df, aes(x = delay_seconds)) +
#   geom_histogram(aes(y = after_stat(density)),
#                  bins = 50, fill = "steelblue", alpha = 0.7, color = "white") +
#   geom_vline(xintercept = 0, linetype = "dashed", color = "coral", linewidth = 1) +
#   coord_cartesian(xlim = c(-300, 600)) +
#   labs(
#     title    = "Delay distribution at {closest_state}",
#     subtitle = "Watch delays shift during morning and evening rush hours",
#     x        = "Delay (seconds)",
#     y        = "Density",
#     caption  = "0s = on time  |  Positive = running late"
#   ) +
#   transition_states(hour_label, transition_length = 2, state_length = 1) +
#   ease_aes("sine-in-out")

# animate(anim, fps = 6, nframes = 72, width = 800, height = 400,
#         renderer = gifski_renderer())
```

---

# Viz 7 — Why Student-t? Heavy tails explained

**What:** Overlay of three density curves — observed delay residuals, the
best-fit Normal, and the fitted Student-t (ν ≈ 2). The shaded region shows
tail probability the Normal model misses.
**Why:** ν = 2.01 means the distribution is near-Cauchy — it has no finite
variance theoretically. Extreme delays (snow storms, accidents) are much more
common than a Normal model would predict.

```{r viz7_plot}
ps        <- posterior_summary(m2)
nu_hat    <- ps["nu",    "Estimate"]
sigma_hat <- ps["sigma", "Estimate"]

x_seq <- seq(-250, 400, by = 0.5)

# Approximate residuals: centre raw delays at their mean
# Compute empirical sd before clipping tails for display so curves match the data
full_r       <- raw6641$delay_seconds - mean(raw6641$delay_seconds)
empirical_sd <- sd(full_r)
resid_approx <- full_r[abs(full_r) <= 250]

curve_df <- bind_rows(
  data.frame(
    x       = x_seq,
    density = dnorm(x_seq, mean = 0, sd = empirical_sd),
    model   = "Normal fit"
  ),
  data.frame(
    x       = x_seq,
    density = dt(x_seq / empirical_sd, df = nu_hat) / empirical_sd,
    model   = sprintf("Student-t (ν = %.1f)", nu_hat)
  )
)

ggplot() +
  geom_histogram(
    data = data.frame(r = resid_approx),
    aes(x = r, y = after_stat(density)),
    bins = 70, fill = "grey75", alpha = 0.6
  ) +
  # Shade tail region (|x| > 100s) to show Normal underestimates
  geom_area(
    data = filter(curve_df, model == "Normal fit", abs(x) > 100),
    aes(x = x, y = density), fill = "steelblue", alpha = 0.25
  ) +
  geom_area(
    data = filter(curve_df, grepl("Student", model), abs(x) > 100),
    aes(x = x, y = density), fill = "coral", alpha = 0.25
  ) +
  geom_line(
    data = curve_df,
    aes(x = x, y = density, color = model, linetype = model),
    linewidth = 1.2
  ) +
  scale_color_manual(
    values = setNames(
      c("steelblue", "coral"),
      c("Normal fit", sprintf("Student-t (ν = %.1f)", nu_hat))
    )
  ) +
  labs(
    title    = "Bus delay distributions have near-Cauchy tails",
    subtitle = sprintf(
      "Fitted ν = %.2f — the Normal drastically underestimates tail probability (shaded)\nRoute 6641  |  σ = %.1fs",
      nu_hat, sigma_hat),
    x = "Delay residual (s)", y = "Density",
    color = NULL, linetype = NULL,
    caption = "Bars = observed delays (centred)  |  Shaded = tail region beyond ±100s"
  ) +
  theme(legend.position = "top")
```

---

# Viz 8 — Bayesian vs frequentist: same conclusion, different language

**What:** The 95% credible interval (Bayesian) vs 95% confidence interval
(frequentist OLS) for the propagation coefficient, side by side.
**Why:** Both agree the slope ≈ 1.0, which validates the finding. But the
Bayesian CI has a direct probability statement: "95% chance the true slope
is in this range given the data." The frequentist CI does not.

```{r viz8_data, cache=TRUE}
lm_fit <- lm(delay_seconds ~ previous_stop_delay + shape_dist_traveled,
             data = train6641)
lm_ci  <- confint(lm_fit)["previous_stop_delay", ]

fe     <- fixef(m2)["previous_stop_delay", ]

ci_df <- data.frame(
  method   = c("Frequentist\n(OLS, 95% CI)", "Bayesian\n(brms M2, 95% CrI)"),
  estimate = c(coef(lm_fit)["previous_stop_delay"], fe["Estimate"]),
  lo       = c(lm_ci["2.5 %"],  fe["Q2.5"]),
  hi       = c(lm_ci["97.5 %"], fe["Q97.5"]),
  stringsAsFactors = FALSE
)
```

```{r viz8_plot, cache=FALSE, fig.height=3.5}
ggplot(ci_df, aes(y = method, x = estimate, xmin = lo, xmax = hi,
                  color = method)) +
  geom_vline(xintercept = 1, linetype = "dashed", alpha = 0.4) +
  geom_errorbarh(height = 0.18, linewidth = 1.5) +
  geom_point(size = 4) +
  scale_color_manual(values = c("steelblue", "coral"), guide = "none") +
  scale_x_continuous(labels = scales::label_number(accuracy = 0.001)) +
  labs(
    title    = "Both methods agree: delay propagation slope ≈ 1.0",
    subtitle = "Bayesian 95% credible interval vs frequentist 95% confidence interval\nDotted line = perfect propagation (slope = 1)",
    x        = "b_previous_stop_delay",
    y        = NULL,
    caption  = sprintf(
      "OLS ignores random effects, splines, Student-t likelihood.\nBayesian R²: %.3f  |  OLS R²: %.3f",
      bayes_R2(m2)[1, "Estimate"], summary(lm_fit)$r.squared
    )
  )
```

---

# Viz 9 — Interactive stop map: where are buses late?

**What:** Every bus stop in the TransLink network colored by mean observed delay.
Green = running early, red = running late. Click a stop for details.
**Why:** The spatial distribution of delays tells a different story than
statistics alone — certain corridors and transfer hubs are consistently late.

```{r viz9_plot, eval=requireNamespace("leaflet", quietly=TRUE), fig.height=6}
library(leaflet)

con_gtfs <- dbConnect(SQLite(), "../database/gtfs_static.db")
stops_geo <- dbGetQuery(con_gtfs,
  "SELECT stop_id, stop_name, stop_lat, stop_lon FROM stops
   WHERE stop_lat IS NOT NULL AND stop_lon IS NOT NULL")
dbDisconnect(con_gtfs)

stop_delays <- rawAll %>%
  group_by(stop_id) %>%
  summarise(
    mean_delay   = mean(delay_seconds, na.rm = TRUE),
    median_delay = median(delay_seconds, na.rm = TRUE),
    n_obs        = n(),
    .groups = "drop"
  ) %>%
  filter(n_obs >= 20)

map_df <- inner_join(stop_delays, stops_geo, by = "stop_id") %>%
  filter(!is.na(stop_lat), !is.na(stop_lon),
         stop_lat > 48.9, stop_lat < 49.6,     # Metro Vancouver bounding box
         stop_lon > -123.5, stop_lon < -122.2)

# Color scale: capped at ±90s for readability
pal <- colorNumeric(
  palette = c("steelblue", "white", "coral"),
  domain  = c(-90, 90),
  na.color = "grey"
)
cap <- function(x) pmax(-90, pmin(90, x))

leaflet(map_df) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(
    lng         = ~stop_lon,
    lat         = ~stop_lat,
    radius      = ~pmax(3, pmin(7, log10(n_obs + 1) * 3)),
    fillColor   = ~pal(cap(mean_delay)),
    fillOpacity = 0.8,
    color       = "white",
    weight      = 0.5,
    popup = ~paste0(
      "<b>", stop_name, "</b><br>",
      "Mean delay: <b>", round(mean_delay, 1), "s</b><br>",
      "Median delay: ", round(median_delay, 1), "s<br>",
      "Observations: ", format(n_obs, big.mark = ",")
    )
  ) %>%
  addLegend(
    "bottomright",
    pal    = pal,
    values = c(-90, 90),
    title  = "Mean delay (s)<br><small>+ = late | − = early</small>",
    labFormat = labelFormat(suffix = "s", transform = function(x) x)
  ) %>%
  setView(lng = -123.12, lat = 49.24, zoom = 11)
```

```{r viz9_fallback, eval=!requireNamespace("leaflet", quietly=TRUE)}
cat("Install leaflet to see the interactive map:\n  install.packages('leaflet')\n")
```
